---
title: "Auditory Evoked M100 Response Latency is Delayed in Children with 16p11.2 Deletion but not 16p11.2 Duplication"
author: "Julian"
date: "July 6, 2015"
output: 
  beamer_presentation: 
    colortheme: dolphin
    fig_crop: no
    keep_tex: yes
    theme: Frankfurt
---

```{r load experimental data, echo = FALSE, eval = TRUE, message = FALSE}
setwd('~/GitHub/reproducible-research/Day-3/sweave_knitr/m100-r-markdown/')
load('~/GitHub/reproducible-research/Day-3/sweave_knitr/m100-r-markdown/child-control-deletion-duplication-analyses-workspace-december-2014.rda')

library(knitr)
```

# Executive summary
- 16p11.2 deletion carriers exhibited delalyed M100 latencies  
- Differences are larger than in 'idiopathic' ASD  
- No oberserved differences between controls and 16p11.2 duplication carriers

# Participants (recruited)
- 137 children  
  + CHOP: 63, UCSF: 74
- 46 16p11.2 deletion carriers  
- 26 16p11.2 duplication carriers  
- 66 age-matched controls  

# Participants (final)
- 99 children  
  + CHOP: 47, UCSF: 52  
- 35 16p11.2 deletion carriers  
- 16 16p11.2 duplication carriers  
- 48 age-matched controls 

# Particpants (final)

|---|Controls	| Deletions	| Duplications|
|-----|-----|-----|-----
|Age range (years) |	7.31 - 17.15	|7.98 - 17.03 |	7.38 - 16.92
|Mean age (years) |	12.82±2.56 |	11.32±.2.65 |	11.45±2.41
|N (evaluable) |	45 |	35 |	16
|NVIQ|	105.73±11.88 |	90.09±15.07 |	83.06±10.68
|VIQ|	107.24±13.72|	85.37±17.05 |	91.69±14.09
|CELF-4|	106.22±11.22 |	75.81±20.80 |	84.62±13.18
|SRS|	16.68±11.89 |	71.51±34.38 |	72.06±41.22
|CTOPP  |	9.23±1.95 |	5.64±2.64 |	7.64±1.45



# Procedure
- passive presentation of binaural sinusoidal tones  
- recording from whole-head magnetometer  
- analysis: artifact rejection, bandpass filtering, source estimation

# Statistical analysis
- M100 latency: dependent variable  
- Linear Mixed Models via `lme4`  
- model: `Hemisphere + Stimulus Condition + Case + Age + Site + (Stimulus Condition + Hemisphere | Subject)`  
- significance: Wald Type II $\chi$^2^ tests; Tukey HSD, correlation tests


# Complete observations
```{r complete observations, echo = FALSE, message = FALSE}
kable(group.complete)
```



# M100 latency effects - condition means

```{r condition means, echo = FALSE, message = TRUE}
library(ggplot2)
cond.means.plot
```

# M100 latency effects - hemisphere means

```{r hemisphere means, echo = FALSE, message = TRUE}
hem.means.plot
```

# M100 latency effects - mean responses
|Mean | Controls | Deletions | Duplications|
|:-----|-----|-----|-----
|Overall (ms) |	127.2±3.1	| 149.0±3.6 |	120.9±5.3
|LH (ms) | 136.7±3.7	| 155.3±4.0 | 126.5±6.6
|RH (ms) | 118.6±3.3 | 143.0±4.1 | 115.7±5.8

# Effect size estimation
```{r case significance Tukey tests, echo = FALSE, message = FALSE}
kable(fortify(child.addmodel.case.tukey), digits = 1)

```

# Variance differences
```{r residuals plus means, echo = FALSE, message = FALSE, warning = FALSE}
residual.plus.mean.density
```

# Conclusion
 - main result: ~20 ms delay in M100 component in 16p11.2 deletion carriers  
 - no observed significant difference bewtween 16p11.2 duplication carriers and controls  
- effects not inconsistent with other reported effects
- support for gene-neurophysiology association